---
title: "lab2"
author: "Lucia Nguyen"
date: "2026-09-07"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

## Task 1
Working directory.
```{r}
getwd()
```

## Task 2
Read and obtain the first six lines of the data.
```{r}
epagas = read.csv("EPAGAS.csv")
head(epagas)
```

## Task 3
- Make the mpg object.Transform mpg to z.
```{r}
mpg = epagas$MPG
z = (mpg - mean(mpg)) / sd(mpg)
head(z)
print(paste0("Mean = ", round(mean(z), digits = 4)), quote = FALSE)
print(paste0("Variance = ", var(z)), quote = FALSE)
print(paste0("Standard Deviation = ", sd(z)), quote = FALSE)
```

- Values of mpg that are possible outliers.
```{r}
mpg[abs(z) >= 2 & abs(z) <= 3]
```

- Values of mpg that defined as outliers.
```{r}
mpg[abs(z) >3]
```

- Construct a dotplot with colors.
```{r}
mpgcol = ifelse(abs(z) > 3, "Red", ifelse(abs(z) >= 2 & abs(z) <= 3, "Blue", "Black"))
library(lattice)
dotplot(mpg, col = mpgcol, main = "Color dotplot for MPG Data")
```

## Task 4
- Boxplot of the mpg variable.
```{r}
boxplot(mpg, main = "MPG Boxplot", xlab = "MPG", col = "black", horizontal = TRUE, notch = TRUE)
```

- Predict with Chebyshev's theorem.

The proportion of data within 2 standard deviations of
the mean, using Chebyshev is at least 
    $1 - \frac{1}{k^2}$
of data lies within k standard deviations. 
For k = 2,
```{r}
k = 2
1 - 1 / (k^2)
```

- Calculate the exact proportion within 2 standard deviation of the mean.
```{r}
exactprop = length(mpg[abs(z) <= 2]) /length(mpg)
exactprop
```

- Does Chebyshev agree with the data?

Yes. Chebyshev's theorem guarantees that at least 75% of the data falls within 2 standard deviations of the mean, and the exact proportion value is 96% -> above that minimum. So the data satisfies the theorem's bound.

- Empirical rule prediction:

The empirical rule states that for roughly bell-shaped data, about 95% of values fall within 2 standard deviations of the mean.
```{r}
   empirical_prop = 0.95
   empirical_prop
```


- How well does it correspond?

Really well. The actual proportion (96%), which is extremely close to the empirical rule's predicted of 95%.

- Is the empirical rule valid in this case? Why?

Yes, it's valid here since the data is roughly bell-shaped/symmetric (in the boxplot and dotplot), and the actual proportion (96%) is very close to what the empirical rule predicted (95%).



